Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-05-16 Manuscript/manuscript.Rmd
manuscript edits

<!--
## HIGH CORRRELATIONS IMPLY COLOCALIZATION OF MANY ANNOTATIONS AROUND TAD BOUNDARIES

By isolating flanked TAD boundaries and, in turn, counting the instances in which annotations overlapped with boundary regions, we were able to assess the correlations of genomic annotations and determine which annotations were enriched together at boundaries. Clustered correlograms for known colocalized annotations showed that distinct groups of transcription factors were highly positively correlated in both cell lines, and across resolutions, including CTCF, ZNF143, RAD21, and SMC3ab (Supplementary Figures 2-9). Likewise, combinations of particular histone modifications and chromatin states were also shown to exhibit similar profiles near TAD boundaries including H3K4me1, H3K4me2, and H3K4me3, as well as strong and weak enhancers. Furthermore, particular chromatin segmentation states, inclduing heterochromatin, repressed chromatin, poised promotors, and repetitive copy number variations, were found to be highly positively inter-correlated, yet were negatively correlated with many transcription factor binding sites. Notably, correlations among annotations were found to increase as resolutions decreased for both cell lines. This indicated that TAD bounardies called, and subsequently flanked, at lower resolutions tended to encapsulate more of these epigenomic marks and contribute to higher levels of enrichment.

The high correlations among many of the genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation. For each model created from the ensemble framework, the alpha and lambda were tuned over a grid of values using 10-fold cross validation (Supplementary Table 7).
-->

@@

In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 5A). We then gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Interestingly, we found that MCCs were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). When including distance-type predictors, models were found to exhibit slightly better MCCs than the other two types (Figure 5C). Importantly, when validating these 3 models built on fully imbalanced datasets, the MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Results were found to be consistently poor across resolutions for the K562 cell line (Supplementary Figure 10). Similarly, many of the performance metrics at higher resolutions were also unable to be reported due to the models inability to produce any true positive cases. <!--Furthermore, we found that evaluating models using conventional metrics, such as AUC or accuracy, did not depict the true severity in performance (Supplementary Table 4).-->

>>>>

## HIGH JACCARD INDICES IMPLY COLOCALIZATION OF MANY GROUP-SPECIFC ANNOTATIONS

Using the jaccard index, we were able to determine which group-specific annotations were enriched together on the linear genome. As could be expected, for GM12878, the class of 14 annotations segmenting the chromation into different states exhibited very low pairwise correlations, with a jaccard index range of (0-0.001) (Table 1). However, specific sets of pairwise histone modifications and transcription factor binding sites (TFBS) were found to be highly enriched. The range of jaccard indices for the 10 histone modifications was found to be (0.022-0.539). Specifically, H3k27ac, H3k4me1, H3k4me2, and H3k4me3 were found to be highly correlated, high levels of which have been detected surrounding transcription start sites [@barski2007high]. For the 79 TFBS, the range of jaccard indices was (<0.001-0.425). Here, known mediators of loop extrusion components were found to have high pairwise correlations including CTCF, RAD21, and SMC3 [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. Similar profiles of jaccard indices were exhibited across the sets of annotations for the K562 cell line (Supplementary Table 4). <!--The high jaccard indices among select sets of genomic annotations gave precedent for the use of an elastic-net regularization algorithm in order to suitably reduce the feature space while retaining possibly correlated annotations that could be predictive of TAD boundary formation.-->

@@

We then went on the evaluate the effect of class imbalance on model performance. In order to set a baseline in performance, we first evaluated the least granular model, that is one implemented using OC predictor types, without data re-sampling. We found that models built on data with fully imbalanced classes exhibited poor MCCs across all resolutions (Figure 5A). We then gradually increased the granularity by assessing models built on OP predictor types, while still not performing any data resampling. Interestingly, we found that MCCs were worse than when using OC predictor types for the GM12878 cell line (Figure 5B). When including distance-type predictors, models were found to exhibit slightly better MCCs than the other two types (Figure 5C). Importantly, when validating these 3 models built on fully imbalanced datasets, the MCC could not be determined at lower resolutions given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives. Results were found to be consistently poor across resolutions for the K562 cell line (Supplementary Figure 10). Similarly, many of the performance metrics at higher resolutions were also unable to be reported due to the models inability to produce any true positive cases. <!--Furthermore, we found that evaluating models using conventional metrics, such as AUC or accuracy, did not depict the true severity in performance (Supplementary Table 4).-->
